\LARGE\textbf{Designing Studies and Evaluating Research Results:\\ Type M and Type S Errors for\\ Pearson Correlation Coefficient }

### 2020-06-01 Documents/Paper_main/Paper_main.Rnw
Final draft
Fix conclusions
Add abstract
Correct test statistic

\begin{abstract}
\end{abstract} \hspace{10pt}

@@

Why do these errors matter? The exaggeration of effect sizes, in the right or in the wrong direction, has important consequences on a theoretical and applied level. On a theoretical level, studies’ designs with high Type M and Type S errors can foster distorted expectations on the effect under study, triggering a vicious cycle for the planning of future studies. This point is relevant also for the design of replication studies, which could turn out to be underpowered if they do not take into account possible inflations of the original effect \parencite{buttonPowerFailureWhy2013}. When studies are used to inform policymaking and real-world interventions, implications can go beyond the academic research community and can impact society at large. In these settings, we could assist to a “hype and disappointment cycle” \parencite{gelmanEmbracingVariationAccepting2019}, where true effects turn out to be much less impressive than expected. This can produce undesirable consequences on people’s lives, a consideration that invites researchers to assume responsibility in effectively communicating the risks related to effects quantification.

@@

In the following sections we illustrate how to perform prospective and retrospective design analysis using some examples. We developed two R functions to perform design analysis for Pearson correlation, which are available at the page \todo{…} . The function to perform a prospective design analysis is \texttt{pro\_r()}. It requires as input the plausible effect size (\texttt{rho}), the statistical power (\texttt{power}) the directionality of the test (\texttt{alternative}) which can be set as: “\texttt{two.sided}”, “\texttt{less}” or “\texttt{greate}r” . Type I error rate (\texttt{sig\_level}) is set as default at 5\% and can be changed by the user. The \texttt{pro\_r()} function returns the necessary sample size to achieve the desired statistical power, Type M error rate, the Type S error probability, and the critical value(s) above which a statistically significant result can be found. The function to perform retrospective design analysis is \texttt{retro\_r()}. It requires as input the plausible effect size, the sample size used in the study, and the directionality of the test that was performed. Also in this case, Type I error rate is set as default at 5\% and can be changed by the user.  The function \texttt{retro\_r()} returns the Type M error rate, the Type S error probability, and the critical value(s). \todo{For further details on (how to use)computational aspects regarding the R functions, we refer to Appendix~\ref{sec:appendix_a}.}

@@

\section{Case Study}

@@

We consider the study published in \emph{Science} by \textcite{eisenbergerDoesRejectionHurt2003} entitled: “Does Rejection Hurt? An fMRI Study of Social Exclusion”. The research question originated from the observation that the Anterior Cingulate Cortex (ACC) is a region of the brain known to be involved in the experience of physical pain. Could pain from social stimuli, such as social exclusion, share similar neural underpinnings? To test this hypothesis, 13 participants were recruited and each one had to play a virtual game with other two players while undergoing functional Magnetic Resonance Imaging (fMRI). The other two players were fictitious, and participants were actually playing against a computer program. Players had to toss a virtual ball among each other in three conditions: social inclusion, explicit social exclusion and implicit social exclusion. In the social inclusion condition the participant regularly received the ball. In the explicit social exclusion condition the participant was told that, due to technical problems, he was not going to play that round. In the implicit social exclusion condition, the participant experienced being intentionally left out from the game by the other two players. At the end of the experiment, each participant completed a self-report measure regarding their perceived distress when they were intentionally left out by the other players. Considering only the implicit social exclusion condition, a correlation coefficient was estimated between the measure of distress and neural activity in the Anterior Cingulate Cortex. As suggested by the large and statistically significant correlation coefficient between perceived distress and activity in the ACC, \todo{r = .88, p < .005} \parencite[p. 291]{eisenbergerDoesRejectionHurt2003}, authors concluded that social and physical pain seem to share similar neural underpinnings.

@@

A retrospective design analysis should start with thorough reasoning on the plausible size and direction of the effect under study. In order to produce valid inferences, a lot of attention should be devoted to this point by integrating external information. For the sake of this example, we turn to the considerations made by \textcite{vulSuspiciouslyHighCorrelations2017} who \todo{considered} correlations between personality measures and neural activity to be likely around $\rho=.25$.  A correlation of $\rho=.50$ was deemed plausible but optimistic and a correlation of $\rho=.75$ was considered theoretically plausible but unrealistic.

@@

\section{Retrospective Design Analysis}

@@

We can use the R function \texttt{retro\_r()}, which inputs and outputs are displayed in Figure~\ref{fig:retro_r}. In this study, the statistical power is .13, that is to say, there is a 13\% probability to reject the null hypothesis, if an effect of at least $\rho=|.25|$ exists. Consider this point together with the results obtained in the experiment: r = .88, p < .005 \parencite[p. 291]{eisenbergerDoesRejectionHurt2003}. It is clear that, even though the probability to reject the null hypothesis is low (power of 13\%), this event could happen. And when it does happen, it is tempting to believe that results are \todo{impressive} \parencite{gelmanStatisticalCrisisScience2014}. However, this design comes with serious inferential risks for the estimation of effect sizes, which could be grasped by presenting Type M and Type S errors. A glance at their value communicates that it is not impossible to find a statistically significant result, but when it does happen, the effect sizes could be largely overestimated - Type M = 2.58 - and maybe even in the wrong direction - Type S = .03. The Type M error rate of \todo{around} 2.60 indicates that a statistically significant correlation is on average about two and a half times the plausible value. In other words, statistically significant results emerging in such a research design will on average overestimate the plausible correlation coefficient by 160\%. The Type S error of  .03 suggests that there is a three percent probability to find a statistically significant result in the opposite direction, in this example, a negative relationship.

@@

\begin{figure}[!h]
\centering
  \includegraphics[height=5cm]{screens/retro}
\caption{Input and Output of the function \texttt{retro\_r()} for retrospective design analysis. Case study: \textcite{eisenbergerDoesRejectionHurt2003}. The plausible correlation coefficient is $\rho = .25$, the sample size is 13, and the statistical test is two-tailed.}\label{fig:retro_r}
\end{figure}

In this research design, the critical values above which a statistically significant result is declared correspond to $\rho=\pm .55$ (Figure~\ref{fig:retro_r}).  These values are highlighted in Figure~\ref{fig:Plot_winner} as the vertical lines in the sampling distribution of correlation coefficients under the null hypothesis. Notice that the plausible effect size lies in the region of acceptance of the null hypothesis. Therefore, it is impossible to simultaneously find a statistically significant result and estimate an effect close to the plausible one ($\rho=.25$). The figure represents the so called Winner’s curse: “the ‘lucky’ scientist who makes a discovery is cursed by finding an inflated estimate of that effect” \parencite{buttonPowerFailureWhy2013}.

@@

<<Plot_winner, fig.pos="!h", fig.align="center", fig.height=3, fig.width=4.5, fig.cap="Winner's course. H$_0$ = Null Hypothesis, H$_1$ = Alternative Hypothesis. When sample size, directionality of the test and Type error probability are set, also the smallest effect size above which is possible to find a statistically significant result is set. In this case, the plausible effect size, $\\rho=.25$, lies in the region where it is not possible to reject H$_0$. Thus, it is impossible to simultaneously find a statistically significant result and an effect close to the plausible one. In other words, a statistically significant effect must exaggerate the plausible effect size.", fig.scap="Winner's course">>=

@@

\section{Prospective Design Analysis}

@@

Imagine that we were part of the research team in the previous case study exploring the relationship between activity in the Anterior Cerebral Cortex and perceived distress. When drafting the research protocol, we face the inevitable discussion on how many participants we are going to recruit. This choice depends on available resources, type of study design, constraints of various nature and, importantly, the plausible magnitude and direction of the phenomenon that we are going to study. As previously mention, deciding on a plausible effect size is a fundamental step which requires great effort and should not be \todo{a game where different numbers are tried until the desired sample size is reached}. Instead, proposing a plausible effect size is where the expert knowledge of the researcher can be formalized and can greatly contribute to the informativeness of the study that is being planned. For the sake of this examples, we adopt the previous consideration and we suppose that common agreement is reached on a plausible correlation coefficient to be around $\rho=.25$. Finally, we would like to leave open the possibility to explore whether the relationship goes in the opposite direction to the one hypothesized, so we decide to perform a two-tailed test.

We can implement the prospective design analysis using the function \texttt{pro\_r()} which inputs and outputs are displayed in Figure~\ref{fig:pro_r}. About 125 participants are necessary to have 80 \% probability to detect an effect of at least $\rho=\pm .25$, if it actually exists. With this sample size, the Type S error is minimized and approximates zero. In this study design, the Type M error is 1.12 indicating that statistically significant results are on average exaggerated by 12\%. It is possible to notice that the critical values are $\rho=\pm .18$, further highlighting that our plausible effect size is actually included among those values that lead to the acceptance of the alternative hypothesis.

@@

\begin{figure}[!h]
\centering
  \includegraphics[height=5cm]{screens/pro}
\caption{Input and Output of the function \texttt{pro\_r()} for prospective design analysis. Plausible correlation coefficient is $\rho = .25$, statistical power is 80\% and the statistical test is two-tailed.}\label{fig:pro_r}
\end{figure}

@@

A lot has been written regarding the automatic adoption of a conventional $\alpha$ level of 5\% \parencite[e.g.,][]{gigerenzerNullRitualWhat2004, lakensJustifyYourAlpha2018}. This practice is increasingly discouraged, and researchers are invited to think about the best trade-off between $\alpha$ level and statistical power, considering the aim of the study and available resources. The $\alpha$ level impacts Type M and Type S errors as much as it impacts statistical power. Everything else equal, Type M error increases with decreasing $\alpha$ level (i.e., negative relationship), whereas Type S error decreases with decreasing $\alpha$ level (i.e., positive relationship). To further illustrate the relation between Type M error and $\alpha$ level, let us take as an example the previous case study with a sample of 13 participants, plausible effect size $\rho=.25$ and two-tailed test. Table~\ref{tab:Table_alpha} shows that by lowering the $\alpha$ level from 10\% to .10\%, the critical values move from \todo{$\rho=\pm .48$  to $\rho=\pm .80$}. This suggests that, with these new higher thresholds, the exaggeration of effects will be even more pronounced because effects have to be even larger to pass such higher critical values. Instead, the relationship between Type S error and $\alpha$ level can be clarified thinking that by lowering the statistical significance threshold, we are being more conservative to falsely reject the null hypothesis in general which implies that we are also being more conservative to falsely rejecting the null hypothesis in the wrong direction.

@@

Another important choice in study design is the directionality of the test (i.e., one-tailed or two-tailed).  Design analysis invites reasoning on the plausible effect size and hypothesizing the direction of the effect, not only its magnitude. So why should a researcher perform nondirectional statistical tests when there is a hypothesized direction? Performing a two-tailed test leaves open the possibility to find an unexpected result in the opposite direction \parencite{cohenStatisticalPowerAnalysis1988}, a possibility which may be of special interest for preliminary exploratory studies. However, in more advanced stages of a research program (i.e., confirmatory study), directional hypotheses benefit from higher statistical power and lower Type M error rates (Figure~\ref{fig:Plot_hypothesis}). As an example, let us consider the differences between a two-tailed test and a one-tailed test in the previous case study. We can perform a new prospective design analysis (Figure~\ref{fig:pro_r2}) with plausible correlation of $\rho=.25$, 80\% statistical power, but this time setting the argument \texttt{alternative} in the R function to \texttt{“greater”}. A comparison of the two prospective design analyses, \todo{Figure~\ref{fig:pro_r} and Figure~\ref{fig:pro_r2}}, suggests that the same Type M error rate of about 10\% is guaranteed with 94 participants, instead of the 125 subjects necessary with a two-tailed test. Note that, Type S error is not possible in directional statistical tests. Indeed, all the statistically significant results are obtainable only in the hypothesized direction, not the opposite one.

@@

\todo{Aggiunta revisione Altoè}

Choices regarding studies’ design impact effect size estimation. Type M (magnitude) error and Type S (sign) error directly quantify these inferential risks. Their consideration in a prospective design analysis increases awareness of what are the consequences of sample size choice reasoning in similar terms to those used in results evaluation. Instead, retrospective design analysis provides further guidance on interpreting research results. More broadly, design analysis reminds researchers that statistical inference should start before data collection and does not end when results are obtained.

\todo{rivedere numerazione paragrafi e sottoparagrafi\\
}

@@

\section{Pearson Correlation and Design Analysis}\label{sec:appendix_a}
\setcounter{figure}{0}
\setcounter{equation}{0}

To conduct a design analysis, it is necessary to know the sampling distribution of the effect of interest not only under the Null-Hypothesis ($H_0$), but also under the alternative Hypothesis ($H_1$). The sampling distribution is the distribution of effects we would observe if \emph{n} observations were sampled over and over again from a population with a given effect. This allows us to evaluate the statistical power and inferential risks of the study considered.

@@

To test an hypothesis or to derive confidence intervals, it is necessary that the sampling distribution of the effect of interest follows a known distribution. In the case of $H_0:\ \rho = 0$, the sample correlation is approximately Normally distributed with Standard Error: $\text{SE}(r) = \sqrt{(1-r^2)/(n-2)}$. Thus, statistical inference is performed considering the \emph{t}-statistic:
\begin{equation}
  t = \frac{r}{\text{SE}(r)} = r \sqrt{\frac{n-2}{1-r^2}},
\end{equation}
that follows a \emph{t}-distribution with $df = n-2$.

However, in the case of $\rho \neq 0$, assuming normality would lead to unreliable results. As we have previously seen, for large values of $\rho$ and small sample sizes the sampling distribution is skewed. To overcome this issue, the Fisher transformation was introduced \parencite{fisherFrequencyDistributionValues1915}:
\begin{equation}
  F(r) = \frac{1}{2}\ ln \frac{1+r}{1-r} = \text{arctanh}(r).
\end{equation}
Applying this transformation, the resulting sample distribution is approximately Normal with $\text{mean} = F(\rho)$ and $\text{SE} = \frac{1}{\sqrt{n-3}}$. Thus, statistical inference is performed considering the \emph{Z}-scores.

Alternatively, other methods can be used to obtain reliable results, as for example Monte Carlo simulation. Monte Carlo simulation is based on random sampling to approximate the quantities of interest. In the case of correlation, \emph{n} observations are iteratively simulated from a bivariate normal distribution with a given $\rho$, and the observed correlation is considered. As the number of iterations increases, the distribution of simulated correlation values approximates the actual correlation sampling distribution and it can be used to compute the quantities of interest.

@@

Design Analysis was first introduced by \textcite{gelmanPowerCalculationsAssessing2014} assuming that the sampling distribution of the effect estimate follows a \emph{t}-distribution. This is the case, for example, of Cohen's \emph{d} effect size. Cohen's \emph{d} is used to measure the mean difference between two groups on a continuous outcome. The behaviour of Type M and Type S errors as a function of statistical power in the case of Cohen's \emph{d} is presented in Figure~\ref{fig:Plot_cohen}.

@@

For different values of hypothetical population effect size \todo{(\emph{d} = .2, .5, .7, .9)}, we can observe that, for high levels of power, Type S and Type M errors are low. Conversely, the Type S and Type M errors are high for low values of power. In particular, the relation between power and inferential errors is not influenced by the value of \emph{d} (i.e., the four lines are overlapping). Limit cases are obtained for power = 1 and 0.05 (note that the lowest value of power is given by the alpha value chosen as the statistical significance threshold). In the former case, Type S error is 0 and Type M error is 1. In the latter case, Type S error is 0.5 and the Type M error value goes to infinity.

In the case of Pearson's Correlation, we noted above that the sampling distribution is skewed for large values of $\rho$ and small sample sizes. Moreover, the support is bounded between -1 and 1. Thus, the relations between power, Type M, and Type S error are influenced by the value of hypothetical population effect size (see Figure~\ref{fig:Plot_derign_rho}).

@@

We can observe how, for different values of correlation \todo{($\rho$ = .2, .5, .7, .9)}, Type M error increases at different rates when the power decrease, whereas Type S error follows a consistent pattern (note that differences are due to numerical approximation). We can intuitively explain this behaviour considering that, for low levels of power, the sampling distribution includes a wider range of correlation values. However, correlation values can not exceed the value 1 and therefore the distribution becomes progressively more skewed. This does not influence the proportion of statistically significant sampled correlations with the incorrect sign (Type S error), but it affects the mean absolute value of statistically significant sampled correlations (used to compute Type M error). In particular, sampling distributions for greater values of $\rho$ becomes skewed more rapidly and thus Type-M error increases at a lower rate.

Finally, since the correlation values are bounded, Type-M error for a given value of $\rho$ can increase only to a maximum value given by $\frac{1}{\rho}$. For example, for $\rho = .5$ the maximum Type-M error is 2 as $.5\times 2 = 1$ (i.e., the maximum correlation value).

>>>>

\begin{abstract}
It is widely appreciated that many studies in psychological science suffer from low statistical power. One of the consequences of analyzing underpowered studies with thresholds of statistical significance, is a high risk of finding exaggerated effect size estimates, in the right or in the wrong direction. These inferential risks can be directly quantified in terms of Type M (magnitude) error and Type S (sign) error, which directly communicate the consequences of design choices on effect size estimation. Given a study design, Type M error is the factor by which a statistically significant effect is on average exaggerated. Type S error is the probability to find a statistically significant result in the opposite direction to the plausible one. Ideally, these errors should be considered during a \emph{prospective design analysis} in the design phase of a study to determine the appropriate sample size. However, they can also be considered when evaluating studies’ results in a \emph{retrospective design analysis}. In the present contribution we aim to facilitate the considerations of these errors in the research practice in psychology. For this reason we illustrate how to consider Type M and Type S errors in a design analysis using one of the most common effect size measures in psychology: Pearson correlation coefficient. We provide various examples and make the R functions freely available to enable researchers to perform design analysis for their research projects.
\end{abstract} \hspace{10pt}

@@

Why do these errors matter? The exaggeration of effect sizes, in the right or in the wrong direction, has important consequences on a theoretical and applied level. On a theoretical level, studies’ designs with high Type M and Type S errors can foster distorted expectations on the effect under study, triggering a vicious cycle for the planning of future studies. This point is relevant also for the design of replication studies, which could turn out to be underpowered if they do not take into account possible inflations of the original effect \parencite{buttonPowerFailureWhy2013}. When studies are used to inform policymaking and real-world interventions, implications can go beyond the academic research community and can impact society at large. In these settings, we could assist to a “hype and disappointment cycle” \parencite{gelmanOverconfidenceResearchCertainty2019}, where true effects turn out to be much less impressive than expected. This can produce undesirable consequences on people’s lives, a consideration that invites researchers to assume responsibility in effectively communicating the risks related to effects quantification.

@@

In the following sections we illustrate how to perform prospective and retrospective design analysis using some examples. We developed two R functions to perform design analysis for Pearson correlation, which are available at the page \todo{…} . The function to perform a prospective design analysis is \texttt{pro\_r()}. It requires as input the plausible effect size (\texttt{rho}), the statistical power (\texttt{power}) the directionality of the test (\texttt{alternative}) which can be set as: “\texttt{two.sided}”, “\texttt{less}” or “\texttt{greate}r” . Type I error rate (\texttt{sig\_level}) is set as default at 5\% and can be changed by the user. The \texttt{pro\_r()} function returns the necessary sample size to achieve the desired statistical power, Type M error rate, the Type S error probability, and the critical value(s) above which a statistically significant result can be found. The function to perform retrospective design analysis is \texttt{retro\_r()}. It requires as input the plausible effect size, the sample size used in the study, and the directionality of the test that was performed. Also in this case, Type I error rate is set as default at 5\% and can be changed by the user.  The function \texttt{retro\_r()} returns the Type M error rate, the Type S error probability, and the critical value(s). For further details regarding the R functions refer to Appendix~\ref{sec:appendix_b}.

@@

\subsection{Case Study}

@@

We consider the study published in \emph{Science} by \textcite{eisenbergerDoesRejectionHurt2003} entitled: “Does Rejection Hurt? An fMRI Study of Social Exclusion”. The research question originated from the observation that the Anterior Cingulate Cortex (ACC) is a region of the brain known to be involved in the experience of physical pain. Could pain from social stimuli, such as social exclusion, share similar neural underpinnings? To test this hypothesis, 13 participants were recruited and each one had to play a virtual game with other two players while undergoing functional Magnetic Resonance Imaging (fMRI). The other two players were fictitious, and participants were actually playing against a computer program. Players had to toss a virtual ball among each other in three conditions: social inclusion, explicit social exclusion and implicit social exclusion. In the social inclusion condition the participant regularly received the ball. In the explicit social exclusion condition the participant was told that, due to technical problems, he was not going to play that round. In the implicit social exclusion condition, the participant experienced being intentionally left out from the game by the other two players. At the end of the experiment, each participant completed a self-report measure regarding their perceived distress when they were intentionally left out by the other players. Considering only the implicit social exclusion condition, a correlation coefficient was estimated between the measure of distress and neural activity in the Anterior Cingulate Cortex. As suggested by the large and statistically significant correlation coefficient between perceived distress and activity in the ACC, r = .88, p < .005 \parencite[p. 291]{eisenbergerDoesRejectionHurt2003}, authors concluded that social and physical pain seem to share similar neural underpinnings.

@@

A retrospective design analysis should start with thorough reasoning on the plausible size and direction of the effect under study. In order to produce valid inferences, a lot of attention should be devoted to this point by integrating external information. For the sake of this example, we turn to the considerations made by \textcite{vulSuspiciouslyHighCorrelations2017} who suggested correlations between personality measures and neural activity to be likely around $\rho=.25$.  A correlation of $\rho=.50$ was deemed plausible but optimistic and a correlation of $\rho=.75$ was considered theoretically plausible but unrealistic.

@@

\subsection{Retrospective Design Analysis}

@@

We can use the R function \texttt{retro\_r()}, which inputs and outputs are displayed in Figure~\ref{fig:retro_r}. In this study, the statistical power is .13, that is to say, there is a 13\% probability to reject the null hypothesis, if an effect of at least $\rho=|.25|$ exists. Consider this point together with the results obtained in the experiment: $r = .88$, $p < .005$ \parencite[p. 291]{eisenbergerDoesRejectionHurt2003}. It is clear that, even though the probability to reject the null hypothesis is low (power of 13\%), this event could happen. And when it does happen, it is tempting to believe that results are even more remarkable \parencite{gelmanStatisticalCrisisScience2014}. However, this design comes with serious inferential risks for the estimation of effect sizes, which could be grasped by presenting Type M and Type S errors. A glance at their value communicates that it is not impossible to find a statistically significant result, but when it does happen, the effect sizes could be largely overestimated - Type M = 2.58 - and maybe even in the wrong direction - Type S = .03. The Type M error rate of  2.58 indicates that a statistically significant correlation is on average about two and a half times the plausible value. In other words, statistically significant results emerging in such a research design will on average overestimate the plausible correlation coefficient by 160\%. The Type S error of  .03 suggests that there is a three percent probability to find a statistically significant result in the opposite direction, in this example, a negative relationship.

@@

\begin{figure}[!h]
\centering
  \includegraphics[height=5cm]{screens/retro}
\caption{Input and Output of the function \texttt{retro\_r()} for retrospective design analysis. Case study: \textcite{eisenbergerDoesRejectionHurt2003}. The plausible correlation coefficient is $\rho = .25$, the sample size is 13, and the statistical test is two-tailed. The option \texttt{seed} allows to set the random number generator to obtain reproducible results.}\label{fig:retro_r}
\end{figure}

In this research design, the critical values above which a statistically significant result is declared correspond to $r=\pm .55$ (Figure~\ref{fig:retro_r}).  These values are highlighted in Figure~\ref{fig:Plot_winner} as the vertical lines in the sampling distribution of correlation coefficients under the null hypothesis. Notice that the plausible effect size lies in the region of acceptance of the null hypothesis. Therefore, it is impossible to simultaneously find a statistically significant result and estimate an effect close to the plausible one ($\rho=.25$). The figure represents the so called Winner’s curse: “the ‘lucky’ scientist who makes a discovery is cursed by finding an inflated estimate of that effect” \parencite{buttonPowerFailureWhy2013}.

@@

<<Plot_winner, fig.pos="!h", fig.align="center", fig.height=3, fig.width=4.5, fig.cap="Winner's course. H$_0$ = Null Hypothesis, H$_1$ = Alternative Hypothesis. When sample size, directionality of the test and Type error probability are set, also the smallest effect size above which is possible to find a statistically significant result is set. In this case, the plausible effect size, $\\rho=.25$, lies in the region where it is not possible to reject H$_0$ (the region delimited by the two vertical lines). Thus, it is impossible to simultaneously find a statistically significant result and an effect close to the plausible one. In other words, a statistically significant effect must exaggerate the plausible effect size.", fig.scap="Winner's course">>=

@@

\subsection{Prospective Design Analysis}

@@

Imagine that we were part of the research team in the previous case study exploring the relationship between activity in the Anterior Cerebral Cortex and perceived distress. When drafting the research protocol, we face the inevitable discussion on how many participants we are going to recruit. This choice depends on available resources, type of study design, constraints of various nature and, importantly, the plausible magnitude and direction of the phenomenon that we are going to study. As previously mention, deciding on a plausible effect size is a fundamental step which requires great effort and should not be done by trying different values only to obtain a more desirable sample size. Instead, proposing a plausible effect size is where the expert knowledge of the researcher can be formalized and can greatly contribute to the informativeness of the study that is being planned. For the sake of this examples, we adopt the previous consideration and we suppose that common agreement is reached on a plausible correlation coefficient to be around $\rho=.25$. Finally, we would like to leave open the possibility to explore whether the relationship goes in the opposite direction to the one hypothesized, so we decide to perform a two-tailed test.

We can implement the prospective design analysis using the function \texttt{pro\_r()} which inputs and outputs are displayed in Figure~\ref{fig:pro_r}. About 125 participants are necessary to have 80 \% probability to detect an effect of at least $\rho=\pm .25$, if it actually exists. With this sample size, the Type S error is minimized and approximates zero. In this study design, the Type M error is 1.12 indicating that statistically significant results are on average exaggerated by 12\%. It is possible to notice that the critical values are $r=\pm .18$, further highlighting that our plausible effect size is actually included among those values that lead to the acceptance of the alternative hypothesis.

@@

\begin{figure}[!h]
\centering
  \includegraphics[height=5cm]{screens/pro}
\caption{Input and Output of the function \texttt{pro\_r()} for prospective design analysis. Plausible correlation coefficient is $\rho = .25$, statistical power is 80\% and the statistical test is two-tailed. The option \texttt{seed} allows to set the random number generator to obtain reproducible results.}\label{fig:pro_r}
\end{figure}

@@

A lot has been written regarding the automatic adoption of a conventional $\alpha$ level of 5\% \parencite[e.g.,][]{gigerenzerNullRitualWhat2004, lakensJustifyYourAlpha2018}. This practice is increasingly discouraged, and researchers are invited to think about the best trade-off between $\alpha$ level and statistical power, considering the aim of the study and available resources. The $\alpha$ level impacts Type M and Type S errors as much as it impacts statistical power. Everything else equal, Type M error increases with decreasing $\alpha$ level (i.e., negative relationship), whereas Type S error decreases with decreasing $\alpha$ level (i.e., positive relationship). To further illustrate the relation between Type M error and $\alpha$ level, let us take as an example the previous case study with a sample of 13 participants, plausible effect size $\rho=.25$ and two-tailed test. Table~\ref{tab:Table_alpha} shows that by lowering the $\alpha$ level from 10\% to .10\%, the critical values move from $r=\pm .48$  to $r=\pm .80$. This suggests that, with these new higher thresholds, the exaggeration of effects will be even more pronounced because effects have to be even larger to pass such higher critical values. Instead, the relationship between Type S error and $\alpha$ level can be clarified thinking that by lowering the statistical significance threshold, we are being more conservative to falsely reject the null hypothesis in general which implies that we are also being more conservative to falsely rejecting the null hypothesis in the wrong direction.

@@

Another important choice in study design is the directionality of the test (i.e., one-tailed or two-tailed).  Design analysis invites reasoning on the plausible effect size and hypothesizing the direction of the effect, not only its magnitude. So why should a researcher perform nondirectional statistical tests when there is a hypothesized direction? Performing a two-tailed test leaves open the possibility to find an unexpected result in the opposite direction \parencite{cohenStatisticalPowerAnalysis1988}, a possibility which may be of special interest for preliminary exploratory studies. However, in more advanced stages of a research program (i.e., confirmatory study), directional hypotheses benefit from higher statistical power and lower Type M error rates (Figure~\ref{fig:Plot_hypothesis}). As an example, let us consider the differences between a two-tailed test and a one-tailed test in the previous case study. We can perform a new prospective design analysis (Figure~\ref{fig:pro_r2}) with plausible correlation of $\rho=.25$, 80\% statistical power, but this time setting the argument \texttt{alternative} in the R function to \texttt{“greater”}. A comparison of the two prospective design analyses, Figure~\ref{fig:pro_r} and Figure~\ref{fig:pro_r2}, suggests that the same Type M error rate of about 10\% is guaranteed with 94 participants, instead of the 125 subjects necessary with a two-tailed test. Note that, Type S error is not possible in directional statistical tests. Indeed, all the statistically significant results are obtainable only in the hypothesized direction, not the opposite one.

@@

Finally, prospective design analysis could contribute to better research design, however many other important factors were not considered in this contribution. For example, the validity and reliability of measurements should be at the forefront in research design, and careful planning of the entire research protocol is of utmost importance. Future works could tackle some of these shortcomings for example, including an analysis of the quality of measurement on the estimates of Type M and Type S errors. Also, we believe that it would be valuable to provide extension of design analysis for other common effect size measures with the development of statistical software packages that could be directly used by researchers. Moreover, design analysis on Pearson correlation can be easily extended to the multivariate case where multiple predictors are considered. Lastly, design analisys is not limited to the Neyman-Pearson framework, but can be considered also within other statistical approaches such as Bayesian approach. Future works could implement design analysis to evaluate the inferential risks related to the use of Bayes Factors and Bayesian Credibility Intervals.

Summarizing, choices regarding studies’ design impact effect size estimation and Type M (magnitude) error and Type S (sign) error allow to directly quantify these inferential risks. Their consideration in a prospective design analysis increases awareness of what are the consequences of sample size choice reasoning in similar terms to those used in results evaluation. Instead, retrospective design analysis provides further guidance on interpreting research results. More broadly, design analysis reminds researchers that statistical inference should start before data collection and does not end when results are obtained.

@@

\section{Pearson Correlation and Design Analysis}\label{sec:appendix_a}
\setcounter{figure}{0}
\setcounter{equation}{0}
\setcounter{footnote}{0}

To conduct a design analysis, it is necessary to know the sampling distribution of the effect of interest. That is, the distribution of effects we would observe if \emph{n} observations were sampled over and over again from a population with a given effect. This allows us, in turns, to evaluate the sampling distribution of the test statistic of interest not only under the Null-Hypothesis ($H_0$), but also under the alternative Hypothesis ($H_1$), and  thus to compute the statistical power and inferential risks of the study considered.

@@

To test an hypothesis or to derive confidence intervals, it is necessary that the sampling distribution of the test statistic of interest follows a known distribution. In the case of $H_0:\ \rho = 0$, the sample correlation is approximately normally distributed with Standard Error: $\text{SE}(r) = \sqrt{(1-r^2)/(n-2)}$. Thus, statistical inference is performed considering the test statistic:
\begin{equation}
  t = \frac{r}{\text{SE}(r)} = r \sqrt{\frac{n-2}{1-r^2}},
\end{equation}
that follows a \emph{t}-distribution with $df = n-2$.

However, in the case of $\rho \neq 0$, the sample correlation is no longer normally distributed. As we have previously seen, the sampling distribution is skewed for large values of $\rho$ and small sample sizes. Thus, the test statistic of interest does not follow a \emph{t}-distribution.\footnote{Note that the \emph{t}-distribution is defined as the distribution of a random variable T where $T = \frac{Z}{\sqrt{V/df}}$. With \emph{Z} a standard Normal, \emph{V} a Chi-squared distribution with \emph{df} the degrees of freedom. Thus, if the sample correlation is not longer approximately normally distributed the test-statistic is not longer \emph{t}-distributed.} To overcome this issue, the Fisher transformation was introduced \parencite{fisherFrequencyDistributionValues1915}:
\begin{equation}
  F(r) = \frac{1}{2}\ ln \frac{1+r}{1-r} = \text{arctanh}(r).
\end{equation}
Applying this transformation, the resulting sampling distribution is approximately Normal with $\text{mean} = F(\rho)$ and $\text{SE} = \frac{1}{\sqrt{n-3}}$. Thus, the test statistic follows a standard Normal distribution and statistical inference is performed considering the \emph{Z}-scores.

Alternatively, other methods can be used to obtain reliable results, as for example Monte Carlo simulation. Monte Carlo simulation is based on random sampling to approximate the quantities of interest. In the case of correlation, \emph{n} observations are iteratively simulated from a bivariate Normal distribution with a given $\rho$, and the observed correlation is considered. As the number of iterations increases, the distribution of simulated correlation values approximates the actual correlation sampling distribution and it can be used to compute the quantities of interest.

@@

Design Analysis was first introduced by \textcite{gelmanPowerCalculationsAssessing2014} assuming that the sampling distribution of the test statistic of interest follows a \emph{t}-distribution. This is the case, for example, of Cohen's \emph{d} effect size. Cohen's \emph{d} is used to measure the mean difference between two groups on a continuous outcome. The behaviour of Type M and Type S errors as a function of statistical power in the case of Cohen's \emph{d} is presented in Figure~\ref{fig:Plot_cohen}.

@@

For different values of hypothetical population effect size (\emph{d} = .2, .5, .7, .9), we can observe that, for high levels of power, Type S and Type M errors are low. Conversely, the Type S and Type M errors are high for low values of power. As expected, the relation between power and inferential errors is not influenced by the value of \emph{d} (i.e., the four lines are overlapping). Limit cases are obtained for power = 1 and 0.05 (note that the lowest value of power is given by the alpha value chosen as the statistical significance threshold). In the former case, Type S error is 0 and Type M error is 1. In the latter case, Type S error is 0.5 and the Type M error value goes to infinity.

In the case of Pearson's Correlation, we noted above that the sampling distribution is skewed for large values of $\rho$ and small sample sizes. Moreover, the support is bounded between -1 and 1. Thus, the relations between power, Type M, and Type S error are influenced by the value of hypothetical population effect size (see Figure~\ref{fig:Plot_rho}).

@@

We can observe how, for different values of correlation ($\rho$ = .2, .5, .7, .9), Type M error increases at different rates when the power decrease, whereas Type S error follows a consistent pattern (note that differences are due to numerical approximation). We can intuitively explain this behaviour considering that, for low levels of power, the sampling distribution includes a wider range of correlation values. However, correlation values can not exceed the value 1 and therefore the distribution becomes progressively more skewed. This does not influence the proportion of statistically significant sampled correlations with the incorrect sign (Type S error), but it affects the mean absolute value of statistically significant sampled correlations (used to compute Type M error). In particular, sampling distributions for greater values of $\rho$ becomes skewed more rapidly and thus Type-M error increases at a lower rate.

Finally, since the correlation values are bounded, Type-M error for a given value of $\rho$ can hypothetically increase only to a maximum value given by $\frac{1}{\rho}$. For example, for $\rho = .5$ the maximum Type-M error is 2 as $.5\times 2 = 1$ (i.e., the maximum correlation value).
